Towards a standardized evaluation of multiple imputation routines

### 2022-03-22 Manuscript/StandardizedEvaluation.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Towards a standardized evaluation of multiple imputation routines},
  pdfauthor={Hanne Oberman and Gerko Vink},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\begin{document}
\maketitle
\begin{abstract}
Developing new imputation methodology has become a very active field.
Unfortunately, there is no consensus on how to perform simulation
studies to evaluate the properties of imputation methods. In this paper
I propose a move towards a standardized evaluation of imputation
methods. To demonstrate the need for standardization, I highlight a set
of potential pitfalls that bring forth a chain of potential problems in
the objective assessment of the performance of imputation routines. This
may lead to suboptimal use of multiple imputation in practice.
Additionally, I suggest a course of action for simulating and evaluating
missing data problems.
\end{abstract}

\hypertarget{to-do}{%
\section{TO DO}\label{to-do}}

\begin{itemize}
\tightlist
\item
  add ADEMP paper (Morris, White, and Crowther 2019), diagnostic
  checking paper (Zhao 2022)
\item
  add PPC (refer to Mingyang, and Gelman \& King)
\item
  expand convergence section
\item
  add RMSE
\item
  add insights from literature review for chapter on missing data in EHR
\item
  add link to simulation script, and \(n\)
\item
  add simulation results
\end{itemize}

@@

Multiple imputation Donald B. Rubin (1987) is a state-of-the-art
technique for drawing valid conclusions from incomplete data, and is
increasingly used in many substantive fields, such as medicine (see e.g.
Pulkki-Råback et al. 2015; Alkire et al. 2015; Stinesen Kollberg et al.
2015), psychology (e.g. Sorbi et al. 2014), economics (e.g. Wong et al.
2015; Dafoe, Oneal, and Russett 2013; Allan and Forder 2015), sociology
(e.g. Crowley and Kolenikov 2014; De Hoon and Van Tubergen 2014), and
official statistics (e.g. Roderick J. Little 2012; Klausch, Hox, and
Schouten 2015).

The technique has earned a permanent spot in research and policy making,
demonstrated e.g.~by the detailed manual created by the National
Research Council (Roderick J. Little et al. 2012). Although a top-down
enforcement of valid ways to handle missing data is not yet very
pronounced, an increasing amount of researchers are embracing multiple
imputation techniques. After all, the principle of multiple imputation
is very intuitive.

The idea behind multiple imputation is to impute (fill in) the missing
values multiple times to obtain a valid estimate of what could have
been. The multiple datasets that are thus obtained can be analyzed by
standard techniques and the analysis results can be combined into a
single inference. In contrast to ad hoc methods for dealing with missing
values (e.g.~list-wise deletion, mean imputation, regression imputation,
last observation carried forward), multiple imputation properly takes
the sources of uncertainty that are related to the missingness problem
into account.

The quality of a solution obtained by multiple imputation depends on the
statistical properties of the incomplete data and the degree to which an
imputation procedure is able to capture these properties when modeling
missing values. In general it holds that modeling missing data becomes
more challenging when the amount of missingness increases. However, when
(strong) relations in the data are present, the observed parts can hold
great predictive power for the models that estimate the missingness. In
that case, multiple imputation would be substantially more efficient
than the ubiquitous list-wise deletion.

When evaluating the statistical properties (and thereby the practical
applicability) of new imputation methodology, researchers most often
make use of simulation studies. In such studies a complete data set is
usually generated from a statistical model, another model is used to
induce missingness, and a set of evaluation criteria is postulated to
evaluate the performance of the imputation method. However, no golden
standard has been established and, as a result, validity of the
simulation and evaluation procedures may differ tremendously from one
developer to another.

The purpose of this paper is threefold: First, I raise some concerns
with respect to evaluating imputation methodology. These concerns stem
from careful consideration with fellow `imputers' and from encounters as
a reviewer for statistical journals. Second, I provide imputation
methodologists with a suggested course of action when simulating missing
data problems. This suggested approach should identify a common ground,
but is in no way intended as an absolute solution. This identifies the
third purpose of this paper: discussion. I hope to elicit critical
thinking with respect to the problems at hand. We are all convinced that
our methodology has some merit. But for sake of progress it would be
much more advantageous if the aim of our evaluations would go beyond
\emph{proving the point} and would legitimately consider the statistical
properties.

\hypertarget{why-some-evaluations-should-not-be-trusted}{%
\subsection{Why some evaluations should not be
trusted}\label{why-some-evaluations-should-not-be-trusted}}

As of today, there is no consensus on how to perform simulation studies
to evaluate the properties of imputation methods. Typically, the
developer of a new imputation routine does some tests by simulations,
but these tests differ across developers. This brings forth a chain of
potential problems in the objective assessment of the performance of
imputation routines that may lead to suboptimal use of multiple
imputation in practice. To demonstrate the broad impact of these
problems, I subdivide the problems in the following three distinct
categories: problems with data generation, problems with missingness
generation and problems with performance evaluation. I further detail
the impact these problems may have on the validity of the performance
evaluation of the imputation routines.

@@

To evaluate the ability of an imputation routine to handle the
non-response model, a form of truth has to be established. Those who
perform simulation studies are in the luxury position to establish the
truth beforehand, either by generating data from a statistical model, or
by sampling data from a sufficiently large observed set. However, when
generating data, two easily overlooked problems may arise.

First, data are often generated following the model (or distribution)
that is also used for imputing the data. The performance of the
imputation approach is then deemed good, which is no surprise as the
evaluated conditions are in favor of the problem that is studied.
Although such procedures may be statistically relevant, the approach
would be no good for even the simplest case. Such simulations pose
serious threats to the applicability of the method as real-life data
hardly ever follow a given theoretical distribution.

Second, data are often generated such that the problem that is being
studied is most pronounced. This results in simulated data that contains
very valuable information structures, i.e.~the correlations between
groups of variables may be very pronounced. In other words, no matter
what type of missingness is induced, the observed parts of the data may
still hold much, if not all, of the information about the missing part
and the performance of the imputation procedure is not surprisingly
evaluated as very good. Also, it may be that the relations in real-life
data are rarely as pronounced as the problem that is being studied.

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\hline
 & $\mu_{Y} = 10$ & $\beta_{X} = 0.4$ & CIW & CR & RMSE$_{p}$ & RMSE$_{c}$\\
\hline
\multicolumn{7}{l}{\textbf{\$n = 50\$}}\\
\hline
\hspace{1em}CCA & 9.961 & 0.341 & 0.823 & 0.94 & 0.866 & NaN\\
\hline
\hspace{1em}Imp. & 10.007 & 0.386 & 0.877 & 0.99 & 0.903 & 1.330\\
\hline
\hspace{1em}True & 10.021 & 0.393 & 0.541 & 0.97 & 0.903 & 0.000\\
\hline
\multicolumn{7}{l}{\textbf{\$n = 500\$}}\\
\hline
\hspace{1em}CCA & 9.941 & 0.343 & 0.232 & 0.85 & 0.897 & NaN\\
\hline
\hspace{1em}Imp. & 9.994 & 0.392 & 0.249 & 0.94 & 0.917 & 1.296\\
\hline
\hspace{1em}True & 9.998 & 0.400 & 0.161 & 0.90 & 0.914 & 0.000\\
\hline
\multicolumn{7}{l}{\textbf{\$n = 5000\$}}\\
\hline
\hspace{1em}CCA & 9.945 & 0.344 & 0.074 & 0.09 & 0.899 & NaN\\
\hline
\hspace{1em}Imp. & 10.000 & 0.397 & 0.078 & 0.91 & 0.916 & 1.298\\
\hline
\hspace{1em}True & 10.002 & 0.398 & 0.051 & 0.95 & 0.917 & 0.000\\
\hline
\end{tabu}

@@

In order to obtain valid imputation inference, the imputation model must
capture the essence of the true non-response mechanism Meng (1994). The
model - if any - that is used to generate the missingness is usually
assumed to be random (MAR) or completely random (MCAR). Missingness
mechanisms that are not random (MNAR) and mechanisms that are considered
non-ignorable see e.g. Donald B. Rubin (1976), are generally ignored
during evaluation of imputation methodologies, except for those methods
that are specifically targeted at non-ignorable applications.

@@

Although one cannot definitively verify if the missingness is random -
after all, for every MNAR mechanism there is a MAR mechanism with equal
fit Molenberghs et al. (2008) - it can be argued that MNAR is the more
likely mechanism for real life missingness scenarios. In such cases, an
indication of the validity of the obtained inference, given that the
assumed missingness mechanism is suspected to be invalid, may be
obtained by performing sensitivity analysis see e.g. Molenberghs et al.
(2014), part 5.

The amount of missingness that is used to simulate the performance of
imputation methods differs between studies. Some studies use only 10\%
missing data where other studies push the limits to additionally
investigate the performance under at least 50\% missing data. This
inconsistent display of simulation results may impact the objectivity of
meta-evaluations over imputation methods, as one method's performance
may appear to be favorable because of the less stringent simulation
conditions. This ultimately may lead to statisticians recommending a
less efficient method to applied researchers, thereby limiting the
efficiency of the imputation approach and unnecessarily lowering the
statistical power.

@@

The evaluation criteria used to assess imputation performance vary from
one developer to another. This is not surprising as people from
different fields could have a different focus on the problem at hand.
However, because multiple imputation was designed as a mode for
statistical inference, a minimal set of statistical properties should at
least be evaluated. I highlight four aspects of evaluation that should
be considered when evaluation imputation routines.

First, developers often only inquire about the `accuracy' (i.e.~how well
can the method reproduce the original data). The goal of multiple
imputation is not to reproduce the data, but to allow for obtaining
valid inference given that the data are incomplete. We are interested in
the correct answer to the research question; not in the truth itself.
This means that, given the framework provided by Donald B. Rubin (1987),
statistical properties such as bias, confidence intervals and the
coverage rate of the confidence intervals should be studied. After all,
the 95\% confidence interval should contain the `true' value at least 95
out of 100 times Neyman (1934), p.~591.

Second, the performance of imputation procedures on distributional
properties is often ignored in simulation studies, and even though the
estimates on the analysis level may be justified, some methods can yield
imputations that may seem completely invalid to applied researchers. For
example, one could very accurately estimate average human height by
filling in negative values and values that are unrealistically large.
While the obtained inference could still be valid under such
imputations, the plausibility of the imputed values given the observed
data should be under scrutiny. Rather, one would prefer an imputation
technique to yield both valid inference and plausible imputations. It
should be studied if an imputation method is prone to deliver such
impractical results, and if so, under what conditions.

Third, sampling variation has always been an essential part of the
evaluation of multiple imputation methodology. However, in order to
obtain information about a method's ability to handle the missing data
problem, or to objectively compare methods on their ability to correct
for missingness, it is not necessary to take sampling variation into
account Vink and van Buuren (2014). After all, we are interested only in
the missing data mechanism, and are not considering the noise induced by
the sampling mechanism for evaluation in such studies.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Towards a standardized evaluation of multiple imputation routines},
  pdfauthor={Hanne Oberman and Gerko Vink},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\usepackage[normalem]{ulem}
% Avoid problems with \sout in headers with hyperref
\pdfstringdefDisableCommands{\renewcommand{\sout}{}}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\begin{document}
\maketitle
\begin{abstract}
Developing new imputation methodology has become a very active field.
Unfortunately, there is no consensus on how to perform simulation
studies to evaluate the properties of imputation methods. In this paper
we propose a move towards a standardized evaluation of imputation
methods. To demonstrate the need for standardization, we highlight a set
of potential pitfalls that bring forth a chain of potential problems in
the objective assessment of the performance of imputation routines. This
may lead to suboptimal use of multiple imputation in practice.
Additionally, we suggest a course of action for simulating and
evaluating missing data problems.
\end{abstract}

@@

Multiple imputation (Donald B. Rubin 1987) is a state-of-the-art
technique for drawing valid conclusions from incomplete data. The
technique has earned a permanent spot in research and policy making,
demonstrated e.g.~by the detailed manual created by the National
Research Council (Roderick J. Little et al. 2012). Although a top-down
enforcement of valid ways to handle missing data is not yet very
pronounced, an increasing amount of researchers are embracing multiple
imputation techniques. After all, the principle of multiple imputation
is very intuitive.

The idea behind multiple imputation is to impute (fill in) the missing
values in incomplete data multiple times, to obtain a valid estimate of
what could have been observed. The multiple datasets that are thus
obtained can be analyzed by standard techniques and the analysis results
can be combined into a single inference. In contrast to ad hoc methods
for dealing with missing values (e.g.~list-wise deletion, mean
imputation, last observation carried forward), multiple imputation
properly takes the sources of uncertainty that are related to the
missingness problem into account. {[}It has been empirically
demonstrated time and again that the inferences on multiply imputed data
yield unbiased and confidence valid results (through Rubin's rules). But
how do we know that this is the case, if we do not have access to the
true but unobserved values of the missing entries? And how can we be
sure that extensions to the methodology yield valid inferences too?
Especially in the field of ML/AI, there are promising new imputation
methods, which may yield even sharper imputations than now-standard
(semi-)parametric imputation methods. That's why we use simulation
studies for the evaluation of imputation routines.{]}

\sout{The quality of a solution obtained by multiple imputation depends
on the statistical properties of the incomplete data and the degree to
which an imputation procedure is able to capture these properties when
modeling missing values. In general it holds that modeling missing data
becomes more challenging when the amount of missingness increases.
However, when (strong) relations in the data are present, the observed
parts can hold great predictive power for the models that estimate the
missingness. In that case, multiple imputation would be substantially
more efficient than the ubiquitous list-wise deletion.}

When evaluating the statistical properties (and thereby the practical
applicability) of imputation methodology, researchers most often make
use of simulation studies. In such studies a complete dataset is usually
generated from a statistical model, another model is used to induce
missingness, {[}then several missing data methods are applied{]} and a
set of evaluation criteria is postulated to evaluate the performance of
the imputation methods. However, no golden standard has been established
to evaluate these imputation routines and, as a result, the validity of
the simulation procedures may differ tremendously from one developer to
another.

The purpose of this paper is threefold: First, to raise some concerns
with respect to evaluating imputation methodology. These concerns stem
from careful consideration with fellow `imputers' and from encounters as
a reviewer for statistical journals. Second, to provide imputation
methodologists with a suggested course of action when simulating missing
data problems. This suggested approach should identify a common ground,
but is in no way intended as an absolute solution. This identifies the
third purpose of this paper: discussion. We hope to elicit critical
thinking with respect to the problems at hand. We are all convinced that
our methodology has some merit. But for sake of progress it would be
much more advantageous if the aim of our evaluations would go beyond
\emph{proving the point} and would legitimately consider the statistical
properties.

\hypertarget{why-some-evaluations-should-not-be-trusted}{%
\section{Why some evaluations should not be
trusted}\label{why-some-evaluations-should-not-be-trusted}}

As of today, there is no consensus on how to perform simulation studies
to evaluate the properties of imputation methods. Typically, the
developer of a new imputation routine does some tests by simulations,
but these tests differ across developers. This brings forth a chain of
potential problems in the objective assessment of the performance of
imputation routines that may lead to sub-optimal use of multiple
imputation in practice. To demonstrate the broad impact of these
problems, we subdivide the problems in the following three distinct
categories: problems with data generation, problems with missingness
generation and problems with performance evaluation. we further detail
the impact these problems may have on the validity of the performance
evaluation of the imputation routines.

@@

To evaluate the ability of an imputation routine to handle \sout{the
non-response model} {[}missing data entries{]}, a form of truth has to
be established. Those who perform simulation studies are in the luxury
position to establish the truth beforehand, by choosing a data
generating mechanism to obtain a complete dataset. Complete data is
typically either generated from a statistical model, or sampled from a
sufficiently large \sout{observed} set {[}model-based vs design-based
simulation{]}. \sout{However, when generating data, two easily
overlooked problems may arise.}

The problem with model-based data generating mechanisms is that method
performance may not translate to empirical data. First, data are often
generated following the model (or distribution) that is also used for
imputing the data. The performance of the imputation approach is then
deemed good, which is no surprise as the evaluated conditions are in
favor of the problem that is studied (Morris, White, and Crowther 2019).
Although such procedures may be statistically relevant, the approach
would be no good for even the simplest empirical case. Such simulations
pose serious threats to the applicability of the method as real-life
data hardly ever follow a given theoretical distribution. {[}Add that
data types may be more messy.{]} Second, data are often generated such
that the problem that is being studied is most pronounced. This results
in simulated data that contains very valuable information structures,
i.e.~the correlations between groups of variables may be very
pronounced. In other words, no matter what type of missingness is
induced, the observed parts of the data may still hold much, if not all,
of the information about the missing part and the performance of the
imputation procedure is not surprisingly evaluated as very good. Also,
it may be that the relations in real-life data are rarely as pronounced
as the problem that is being studied.

The problem with design-based simulation is that it's difficult to find
large datasets without missing entries. And imputing the missing data
beforehand and then treating the once-imputed data as comparative truth
may favor the used imputation method in further evaluations. A second
issue is that SEs may be incorrect if the (re)sampling is not handled
correctly. {[}Add pooling when the sample is the population (Vink and
van Buuren 2014) and write something about bootstrapping SEs{]}.

Problems with either of these data generation approaches is that the
number of observations, the number of variables, the variable type(s)
and the coherence between variables may give some imputation methods a
comparative advantage, if they are not a) varied across simulation
conditions {[}when is this fair? is this even an option?{]}, or b) the
object of the study {[}add some examples of simulations specifically
focused on small samples, categorical data, etc.{]}.

{[}Maybe add that some simulation studies do not use a `ground truth' at
all, they just look for the method with the best predictive performance
for a certain completely observed target variable. I don't know where to
introduce the difference between statistical inference and prediction.
But this is an important third category, it seems. These studies apply
different combinations of imputation and estimation methods on one or
more benchmark datasets, and assess the RMSE of the predicted values
after imputation and estimation to say something about the comparative
performance of the methods. In this context, there is no talk of
retrieving the `true' but unobserved values at all. Refer to missing
data chapter (Liu et al. 2021).{]}

@@

Missingness comes in many shapes and forms {[}platitude, but how else to
start?{]}. The most obvious characteristics of missing data are the
missingness pattern(s) and missingness mechanism(s). Missingness
patterns concern the location of missing entries across variables in an
incomplete dataset, whereas missingness mechanisms describe the
relationship between missingness and the values of variables in the data
(Little and Rubin 2020, 8).

The missing data pattern may inadvertently guide which imputation method
prevails in simulation studies. Some ad hoc methods are known to yield
valid inferences in very specific missing data patterns (e.g., list-wise
deletion when there is missingness in the outcome variable of the
analysis model exclusively), while they bias inferences in any other
case. Extrapolating simulation results from such a specific missing data
pattern to more intricate empirical missingness patterns could lead to
false conclusions {[}re-read this! and add monotone/univariate vs
multivariate and sporadic vs systematic dichotomies{]}. Related to the
missingness pattern across variables is the missingness proportion. The
terminology on this topic can be confusing. Does a missingness
proportion of 50\% mean that half of the entries in an incomplete
dataset are missing, or that half of the rows have at least one missing
entry? We will refer to the latter as the proportion of incomplete
cases. But this is not a clear distinction in the literature, which may
lead to incorrect conclusions {[}re-phrase this too{]}. Moreover, the
amount of missingness that is used to simulate the performance of
imputation methods differs between studies. {[}Add ref to chapter (Liu
et al. 2021) and add info.{]} Some studies use only 10\% missing data
where other studies push the limits to additionally investigate the
performance under at least 50\% missing data. This inconsistent display
of simulation results may impact the objectivity of meta-evaluations
over imputation methods, as one method's performance may appear to be
favorable because of the less stringent simulation conditions. This
ultimately may lead to statisticians recommending a less efficient
method to applied researchers, thereby limiting the efficiency of the
imputation approach and unnecessarily lowering the statistical power.

{[}Briefly introduce missingness mechanisms here and show potential
problems, but without simulation results (moved to separate section).
This part should contain: 1) definitions of mechanisms, 2) that MCAR is
often ignored, while it's the minimum requirement and sometimes
realistic in practice, 3) that MAR can be generated \emph{spuriously} if
the correlation between variables is low, and 4) that MNAR is often
ignored, while it may be the most realistic in practice. And finally
something about types of M(N)AR, but what are specific problems with
that?{]}

\hypertarget{section}{%
\section{}\label{section}}

\hrule

\hypertarget{section-1}{%
\section{}\label{section-1}}

{[}Until next hline: to be edited!{]} In order to obtain valid
imputation inference, the imputation model must capture the essence of
the true non-response mechanism Meng (1994). The model - if any - that
is used to generate the missingness is usually assumed to be random
(MAR) or completely random (MCAR). Missingness mechanisms that are not
random (MNAR) and mechanisms that are considered non-ignorable see e.g.
Donald B. Rubin (1976), are generally ignored during evaluation of
imputation methodologies, except for those methods that are specifically
targeted at non-ignorable applications.

@@

Although one cannot definitively verify if the missingness is random -
after all, for every MNAR mechanism there is a MAR mechanism with equal
fit (Molenberghs et al. 2008) - it can be argued that MNAR is the more
likely mechanism for real life missingness scenarios. In such cases, an
indication of the validity of the obtained inference, given that the
assumed missingness mechanism is suspected to be invalid, may be
obtained by performing sensitivity analysis see e.g. Molenberghs et al.
(2014), part 5.

\hypertarget{section-2}{%
\section{}\label{section-2}}

\hrule

@@

{[}Add bridge from last section. And explain how this part is
subdivided: first general characteristics of the imputations such as
convergence, then performance measures.{]}

{[}Performance of different imputation methods may be influenced by
preconditions of the methods themselves. All imputation models should be
congenial, but this is difficult to test to often ignored even in single
imputation endeavor, let alone in simulation studies with 100s of
imputations. Same goes for convergence: FCS is iterative and requires
algorithmic convergence, but this is typically evaluated through visual
inspection which is unfeasible in simulation studies. Ignoring
congeniality may yield sub-optimal imputations, which may disadvantage
certain methods compared to others.{]}

Moreover, the performance of imputation procedures on distributional
properties is often ignored in simulation studies, and even though the
estimates on the analysis level may be justified, some methods can yield
imputations that may seem completely invalid to applied researchers. For
example, one could very accurately estimate average human height by
filling in negative values and values that are unrealistically large.
While the obtained inference could still be valid under such
imputations, the plausibility of the imputed values given the observed
data should be under scrutiny. Rather, one would prefer an imputation
technique to yield both valid inference and plausible imputations. It
should be studied if an imputation method is prone to deliver such
impractical results, and if so, under what conditions.

{[}Then onto the performance measure part.{]}

The evaluation criteria used to assess imputation performance vary from
one developer to another. This is not surprising as people from
different fields could have a different focus on the problem at hand.
But the choice of performance measure(s) may inadvertently distort
statistical properties of the imputed data.

\sout{However, because multiple imputation was designed as a mode for
statistical inference, a minimal set of statistical properties should at
least be evaluated. We highlight four aspects of evaluation that should
be considered when evaluation imputation routines.}

Developers often only inquire about the `accuracy' (i.e.~how well can
the method reproduce the original data). {[}Add RMSE at cell level
here?{]} The goal of multiple imputation is not to reproduce the data,
but to allow for obtaining valid inference given that the data are
incomplete. We are interested in the correct answer to the research
question; not in the truth itself. This means that, given the framework
provided by Donald B. Rubin (1987), statistical properties such as bias,
confidence intervals and the coverage rate of the confidence intervals
should be studied. After all, the 95\% confidence interval should
contain the `true' value at least 95 out of 100 times Neyman (1934),
p.~591. {[}So, what could go wrong by focusing on accuracy
exclusively?{]}

{[}Add choice of analysis model/estimand here, since performance measure
are defined wrt the estimand, e.g., focusing on univariate estimates may
favor invalid ad hoc methods.{]}

{[}Add problems with RMSE here.{]}

Sampling variation has always been an essential part of the evaluation
of multiple imputation methodology. However, in order to obtain
information about a method's ability to handle the missing data problem,
or to objectively compare methods on their ability to correct for
missingness, it is not necessary to take sampling variation into account
Vink and van Buuren (2014). After all, we are interested only in the
missing data mechanism, and are not considering the noise induced by the
sampling mechanism for evaluation in such studies.
